---
title: "Part2_data_analysis"
author: "Qi Li"
date: "2025-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(car)
```

## Data loading and set_up

```{r cars}
# Load your specific cleaned file
# We use read.csv to ensure it loads as a standard dataframe
data_raw <- read.csv("hour_cleaned_1.csv")

# CRITICAL STEP: Convert categorical columns to Factors
# Even though the CSV has "1, 2, 3", R sees these as numbers (1 < 2 < 3).
# We must tell R they are categories so it treats them as separate groups.
#data_raw$season     <- as.factor(data_raw$season)
#data_raw$yr         <- as.factor(data_raw$yr)
#data_raw$weekday    <- as.factor(data_raw$weekday)
#data_raw$holiday    <- as.factor(data_raw$holiday)
#data_raw$weathersit <- as.factor(data_raw$weathersit)
```

## 2. IMPROVEMENT & TRANSFORMATION

# Improvement 1: Cyclical Transformation for Hour
To turn the numbers 0–23 into a circle, we use trigonometry. We convert the "Hour" into coordinates $(x, y)$ on a unit circle.Convert to Radians:First, we take the hour $hr$ and divide it by 24 to get the percentage of the day that has passed. We multiply by $2\pi$ (which represents a full circle in geometry).
```{r pressure, echo=FALSE}
# Filter strict zeros only where necessary for math (Log of 0 is -Infinity)
hour_cleaned_p <- data_raw %>%
  filter(temp > 0, windspeed > 0, hum > 0)

# Improvement 1: Cyclical Transformation for Hour
# This fixes the "Midnight Problem" by connecting Hour 23 back to Hour 0 using Sine/Cosine
hour_cleaned_p$hr_sin <- sin(2 * pi * hour_cleaned_p$hr / 24)
hour_cleaned_p$hr_cos <- cos(2 * pi * hour_cleaned_p$hr / 24)

```

# Improvement 2: Improvement 2: Power Transformations (from your Box-Cox analysis)
```{r}
model_p <- lm(cnt ~ season + weekday + holiday + yr +
                   temp + windspeed + weathersit + hr + hum
             + season * temp, data = hour_cleaned_p)

powerTransform(model_p)
p <- powerTransform(cbind(hour_cleaned_p[,c(5,6,9,10)]))
summary(p)
```

We apply boxcox information to the filtered dataset
```{r}
# cnt: 0.33 -> Cube root
hour_cleaned_p$bccnt <- (hour_cleaned_p$cnt)^(1/3)

# temp: 0.81 -> 0.8
hour_cleaned_p$bctemp <- (hour_cleaned_p$temp)^(0.8)

# hum: 0.89 -> 0.9
hour_cleaned_p$bchum <- (hour_cleaned_p$hum)^(0.9)

# windspeed: -0.02 -> Log
hour_cleaned_p$bcwindspeed <- log(hour_cleaned_p$windspeed)
```

## 3. MODEL REFINEMENT (INTERACTIONS & SENSITIVITY)

And we refit the model again to Calculate Cook's Distance.

# Improvement 3: Added 'weekday * hour' interaction to capture different 
Interaction Terms By adding weekday * (hr_sin + hr_cos), you tell the model: "Do not force the daily curve to look the same. Calculate a separate curve shape for Weekdays and a separate curve shape for Weekends."

```{r}
# peak times for commuters (Mon-Fri) vs leisure riders (Sat-Sun).
model_bc <- lm(bccnt ~ yr + season + holiday + weathersit + 
                    bctemp + bcwindspeed + bchum + 
                    hr_sin + hr_cos +
                    season:bctemp + 
                    weekday * (hr_sin + hr_cos), 
                  data = hour_cleaned_p)
```

# Improvement 4: Remove Influential Points (Sensitivity Analysis)

```{r}
# Calculate Statistics
n <- nrow(hour_cleaned_p)
p <- length(coef(model_bc)) - 1

# 1. COOK'S DISTANCE (Global Influence)
D_i <- cooks.distance(model_bc)
# CHANGE: Use 4/n instead of qf() to actually find points in this large dataset
cookcut <- 4 / n 

# 2. LEVERAGE (Extreme Predictor Values)
h_ii <- hatvalues(model_bc)
hcut <- 2 * ((p + 1) / n)

# 3. DFFITS (Influence on Prediction)
dffits_i <- dffits(model_bc)
fitcut <- 2 * sqrt((p + 1) / n)

plot(model_bc, which = 4, main = "Cook's Distance Plot")
abline(h = cookcut, col = "red", lty = 2)
```

```{r}
# Identify influential points using Cook's Distance (The most robust metric)
influential_indices <- which(D_i > cookcut)

cat("Original Data Size:", n, "\n")
cat("Influential Points Identified:", length(influential_indices), "\n")

# Create "Clean" Data
# We remove the points that fail the Cook's Distance test
clean_data <- hour_cleaned_p[-influential_indices, ]
```

```{r}
model_final <- lm(bccnt ~ yr + season + holiday + weathersit + 
                    bctemp + bcwindspeed + bchum + 
                    hr_sin + hr_cos +
                    season:bctemp + 
                    weekday * (hr_sin + hr_cos), 
                  data = clean_data)

summary(model_final)

# Diagnostic Plot: Check if the red line is flatter (better linearity)
par(mfrow = c(2, 2))
plot(model_final, id.n = 0)

# Normality Check: The dots should be closer to the line now
qqnorm(resid(model_final))
qqline(resid(model_final), col = "red", lwd = 2)

# Print how many influential points were removed
removed_count <- nrow(hour_cleaned_p) - nrow(model_final)
cat("Removed", removed_count, "influential points to improve stability.")
```

```{r}
vif(model_final)
```

```{r}
n <- nrow(clean_data)
p <- length(coef(model_final)) - 1

# leverage cutoff
hcut <- 2 * ((p + 1) / n)

# cook's distance cutoff
cookcut <- qf(0.5, p + 1, n - p - 1)

# dffits cutoff
fitcut <- 2 * sqrt((p + 1) / n)

# dfbeta cutoff
betacut <- 2 / sqrt(n)

# leverage statistic
h_ii <- hatvalues(model_final)

# standardized residuals
r_i <- rstandard(model_final)

# cook's distance 
D_i <- cooks.distance(model_final)

# dffits
dffits_i <- dffits(model_final)

# dfbetas
dfbetas_i <- dfbetas(model_final)

# which observations are leverage points?
which(h_ii > hcut)

# which observations are regression outliers?
which(r_i > 4 | r_i < -4)

# which observations are influential by cook's distance?
which(D_i > cookcut)

# which observations are influential by dffits?
which(abs(dffits_i) > fitcut)

# which observations are influential by dfbetas?
for(i in 1:8){
  print(paste0("Beta ", i-1))
  print(which(abs(dfbetas_i[,i]) > betacut))    # this checks all betas in a loop
}
```
